---
title: "Historic population samples reveal ongoing adaptations of malaria parasites to their human hosts"
author: "Fadel"
date: "`r Sys.Date()`"
output: html_document
---

# 📂 Suggested Folder Structure

```
project_root/
│
├── data/                      # Raw & intermediate data
│   ├── raw/                   # Unprocessed input VCFs & metadata
│   ├── processed/             # Filtered/cleaned datasets
│   ├── allele_freq/           # Allele frequency outputs
│   └── references/            # Reference files (positions, samples, gene lists)
│
├── scripts/                   # Standalone R and bash scripts
│   ├── 01_filter_missing_genotypes.R
│   ├── 02_estimate_Fst.R
│   ├── utils_functions.R      # Reusable R functions
│   └── bash/                  # Bash helper scripts
│
├── results/                   # Outputs
│   ├── figures/               # Plots (PDF, PNG)
│   ├── tables/                # Processed tables, summary stats
│   └── reports/               # Final results
│
├── docs/                      # Documentation
│   └── project_notebook.Rmd   # Main RMarkdown analysis file
│
└── scripts/codes/             # External scripts (e.g., perl, vcflib)
```


# 1. Project Overview

This study compares drug resistance markers from historic *Plasmodium falciparum* samples (1966–72) and recent samples (2015) collected in The Gambia.  

- Historic samples: `1136-PF-GM-NGWA_genotyped`
- Recent samples: `Pf_1162_GM` and `Pf_1182_GM`

**Objectives:**
1. Neutrality tests: Tajima’s D, iHS, BetaScores  
2. Population differentiation: wcFST, Gst, Jost’s D  
3. Allele frequency trajectories across time points (1966, 2001, 2008, 2015)  

---

# 2. Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(data.table)
library(VariantAnnotation)
````

Define paths:

```{r paths}
raw_dir      <- "data/raw"
proc_dir     <- "data/processed"
freq_dir     <- "data/allele_freq"
ref_dir      <- "data/references"
results_dir  <- "results"

scripts_dir  <- "scripts"
```

---

# 3. Data Preparation

## 3.1 Historic Samples (1966–72)

```{bash}
VCF="data/raw/1136-PF-GM-NGWA_genotyped.vcf.gz"
bcftools query -f '%CHROM\t%POS\n' $VCF > $ref_dir/chromosomes.txt
```

Filter non-core chromosomes and low-quality variants:

```{r}
read_delim(file.path(ref_dir, "chromosomes.txt"), col_names = FALSE) %>%
  filter(X1 == "Pf3D7_API_v3") %>%
  write.table(file.path(ref_dir, "chromosomestoRemove.txt"), 
              col.names = FALSE, row.names = FALSE, quote = FALSE)
```

```{bash}
QUAL=10
MIN_DEPTH=5
OUT="data/processed/gam1966_72_biallelic"

vcftools --gzvcf $VCF \
  --exclude-positions $ref_dir/chromosomestoRemove.txt \
  --minQ $QUAL --min-meanDP $MIN_DEPTH --minDP $MIN_DEPTH \
  --max-alleles 2 --min-alleles 2 \
  --recode --recode-INFO-all --out $OUT

mv $OUT.recode.vcf $OUT.vcf
```

Clean missing genotypes:

```{r}
source(file.path(scripts_dir, "01_filter_missing_genotypes.R"))
haps1966 <- filter_missing(file.path(proc_dir, "gam1966_72_biallelic.vcf"))
```

---

## 3.2 Recent Samples (2015)

Merge and filter:

```{bash}
bcftools merge -Oz -o data/processed/Pf_1162_1182_GM.vcf.gz \
  data/raw/1162-PF-GM-NGWA_genotyped.vcf.gz \
  data/raw/1182-PF-GM-DALESSANDRO-SM_genotyped.vcf.gz

tabix -f data/processed/Pf_1162_1182_GM.vcf.gz
```

---

# 4. Neutrality Tests

* Tajima’s D
* iHS
* BetaScores

(See scripts in `scripts/neutrality/` for implementation details.)

---

# 5. Population Differentiation

```{r}
source(file.path(scripts_dir, "02_estimate_Fst.R"))
fst_results <- estimate_Fst(vcf_1966, vcf_2015)
write_tsv(fst_results, file.path(results_dir, "tables/Fst_summary.tsv"))
```

---

# 6. Allele Frequencies Across Time

Prepare allele frequency files:

```{bash}
vcftools --vcf data/processed/gam1966_72.vcf \
  --keep $ref_dir/samples1966.txt \
  --positions $freq_dir/nsSNPspositions.txt \
  --freq2 --out $freq_dir/pop1966
```

Repeat for 2001, 2008, 2015 datasets.

---

# 7. Visualization

## 7.1 Allele Frequency Trajectories

```{r}
freq_combined <- read_tsv(file.path(freq_dir, "allele_frequencies_overTime.tsv"))
ggplot(freq_combined, aes(x = time, y = ref, group = interaction(chrom,pos), color = chrom)) +
  geom_line() + geom_point() +
  labs(title = "Trajectory of Allele Frequencies", x = "Year", y = "Allele Frequency") +
  theme_minimal()
```

## 7.2 Neutrality & Selection Scans

* Manhattan plots for iHS, BetaScores
* Highlight drug resistance loci (PfCRT, DHFR, DHPS, PfAAT1)

---

# 8. Documentation

* All raw input files are in `data/raw/`
* Each intermediate VCF is stored in `data/processed/`
* Scripts for filtering & analysis are in `scripts/`
* Final results, plots, and tables are in `results/`
* This RMarkdown serves as the **master notebook** to reproduce the study.

```

---

This way:  
- The **RMarkdown** acts as the master narrative of your analysis.  
- **Reusable scripts** handle tasks like filtering, Fst estimation, etc.  
- **Folder structure** keeps raw, processed, and results data separated.  
- Each step is fully documented and reproducible.  

---

👉 Do you want me to **populate the standalone scripts** (e.g., `01_filter_missing_genotypes.R` and `02_estimate_Fst.R`) with full working functions so you can directly drop them into your project, or just leave placeholders?
```
